% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrreprt}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother


% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}



\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}



 


\usepackage{tcolorbox}
\usepackage{amssymb}
\usepackage{yfonts}
\usepackage{bm}


\newtcolorbox{greybox}{
  colback=white,
  colframe=blue,
  coltext=black,
  boxsep=5pt,
  arc=4pt}
  
\newcommand{\sectionbreak}{\clearpage}

 
\newcommand{\ds}[4]{\sum_{{#1}=1}^{#3}\sum_{{#2}=1}^{#4}}
\newcommand{\us}[3]{\mathop{\sum\sum}_{1\leq{#2}<{#1}\leq{#3}}}

\newcommand{\ol}[1]{\overline{#1}}
\newcommand{\ul}[1]{\underline{#1}}

\newcommand{\amin}[1]{\mathop{\text{argmin}}_{#1}}
\newcommand{\amax}[1]{\mathop{\text{argmax}}_{#1}}

\newcommand{\ci}{\perp\!\!\!\perp}

\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\mb}[1]{\mathbb{#1}}
\newcommand{\mf}[1]{\mathfrak{#1}}

\newcommand{\eps}{\epsilon}
\newcommand{\lbd}{\lambda}
\newcommand{\alp}{\alpha}
\newcommand{\df}{=:}
\newcommand{\am}[1]{\mathop{\text{argmin}}_{#1}}
\newcommand{\ls}[2]{\mathop{\sum\sum}_{#1}^{#2}}
\newcommand{\ijs}{\mathop{\sum\sum}_{1\leq i<j\leq n}}
\newcommand{\jis}{\mathop{\sum\sum}_{1\leq j<i\leq n}}
\newcommand{\sij}{\sum_{i=1}^n\sum_{j=1}^n}
	
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{bookmark}{}{\usepackage{bookmark}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Gifi Analysis of Multivariate Data},
  pdfauthor={Jan de Leeuw},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}


\title{Gifi Analysis of Multivariate Data}
\author{Jan de Leeuw}
\date{May 5, 2025}
\begin{document}
\maketitle

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}

\bookmarksetup{startatroot}

\chapter*{Preface}\label{preface}
\addcontentsline{toc}{chapter}{Preface}

\markboth{Preface}{Preface}

In 1980 members of the Department of Data Theory at the University of
Leiden taught a post-doctoral course in Nonlinear Multivariate Analysis.
The course content was sort-of-published, in Dutch, as Gifi
(\citeproc{ref-gifi_B_80}{1980}). The course was repeated in 1981, and
this time the sort-of-published version (Gifi
(\citeproc{ref-gifi_B_81}{1981})) was in English.

The preface gives some details about the author.

\begin{quote}
The text is the joint product of the members of the Department of Data
Theory of the Faculty of Social Sciences, University of Leiden. `Albert
Gifi' is their `nom de plume'. The portrait, however, of Albert Gifi
shown here, is that of the real Albert Gifi to whose memory this book is
dedicated, as a far too late recompense for his loyalty and devotion,
during so any years, to the Cause he served.
\end{quote}

Roughly ten years later a revised version of these course notes came out
as an actual book in the \emph{Wiley Series in Probabilty and
Mathematical Statistics} (Gifi (\citeproc{ref-gifi_B_90}{1990})). This
despite the fact that the contents of the book had very little to do
with either probability or mathematical statistics. The book is
organized around a series of computer programs for correspondence
analysis, principal component analysis, and canonical analysis. The
programs, written in FORTRAN, are called HOMALS, PRINCALS, PRIMALS,
CRIMINALS, CANALS, OVERALS because they combine classical linear
multivariate analysis with optimal transformation of the variables,
using alternating least squares (or ALS). It serves, to some extent, as
a manual for the programs, but it also discusses the properties of the
techniques implemented in the programs, and it presents many detailed
applications of these techniques.

Reviewers generally had some difficulties separating the wheat from the
chaff.

\begin{quote}
As the spirit of Albert Gifi has faded away, so has his whimsical
approach to publishing, and his latest book is an idiosyncratic account
of multivariate methods developed by the Leiden group during the 1970s.
The names of their computer programs are distinguished by the ending
\textasciitilde ALS, thus we have OVERALS, PRINCALS, HOMALS, CANALS,
MORALS, MANOVALS, CRIMINALS, PARTALS and PATHALS. Perhaps if you have a
warped mind like this reviewer, you will turn rapidly to CRIMINALS. What
can it be ? Surely it must give some illicit view of the truth about the
world, a vision of the underworld of multivariate analysis ? Alas no !
It turns out only to be a synonym of Canonical Variate Analysis,
sometimes known as Multiple Discriminant Analysis. Likewise HOMALS turns
out to be Reciprocal Averaging, otherwise known as Correspondence
Analysis. (Hill (\citeproc{ref-hill_90}{1990}))
\end{quote}

This ambiguity and confusion are not too surprising. The Gifi book was a
summary of the work of a large number of people, over a period of almost
20 years. Nevertheless, and perhaps because of this, it is somewhat of a
\emph{camel}, which we define for our purposes as a \emph{horse designed
by a committee}. Different chapters had different authors, and the
common ideas behind the various techniques were not always clearly
explained.

\begin{quote}
In Gifi's MVA the criterion called ``meet'' loss plays a central role.
Although the adoption of this criterion is one of the most important
contributions of Gifi, the book would have been much more readable if
this criterion had been introduced right at the outset and was followed
throughout the rest of the book. (Takane
(\citeproc{ref-takane_92}{1992}))
\end{quote}

Nevertheless there is much original material in Gifi
(\citeproc{ref-gifi_B_90}{1990}), and the book has early applications of
alternating least squares, majorization, coordinate descent, the delta
method, and the bootstrap. And it emphasizes throughout the idea that
statistics is about techniques, not about models. But, yes, the
organization leaves much to be desired. An on demand printing of the
first and only edition is now available on Amazon for \$ 492 -- although
of course used versions go for much less.

The book was published by a prestiguous publisher in a prestiguous
series, but it is fair to say it never really caught on. It is not hard
to understand why. The content, and the style, are unfamiliar to
statisticians and mathematicians. There is no inference, no probability,
and very little rigor. The content is in multivariate data analysis,
which would be most at home these days, if anywhere, in a computer
science department. The Gifi group did not have the resources of, say,
Benzécri in France or Hayashi in Japan. The members were mostly active
in psychometrics, a small and insular field, and they were from The
Netherlands, a small country prone to overestimate its importance
(Marvell (\citeproc{ref-marvell_53}{1653})). They also did not have the
evangelical zeal necessary for creating and sustaining a large impact.

There have been some other major publication events in the Gifi saga.
Around the same time as the Wiley book there was the publication of SPSS
(\citeproc{ref-spss_89}{1989}). Starting in the late seventies the Gifi
FORTRAN programs had been embedded in the SPSS system. The \emph{SPSS
Categories} manual was updated many times, in fact every time SPSS or
IBM SPSS had a new release. Over the years other programs produced by
the Department of Data Theory were added. A recent version is, for
example, Meulman and Heiser (\citeproc{ref-meulman_heiser_12}{2012}),
corresponding to IBM SPSS 21. It acknowledges the contributions of some
of the members of the Gifi team -- but in IBM
(\citeproc{ref-ibm_15}{2015}), the version for IBM SPSS 23, these
acknowledgements and the names of the authors have disappeared. Sic
transit gloria mundi.

Michailidis and De Leeuw (\citeproc{ref-michailidis_deleeuw_A_98}{1998})
made an attempt to make the Gifi material somewhat more accessible by
publishing a review article in a widely read mainstream statistical
journal. Another such attempt is De Leeuw and Mair
(\citeproc{ref-deleeuw_mair_A_09a}{2009}), in which the homals package
for R is introduced. The homals package is basically a single monolithic
R function that can do everything the Gifi programs can do, and then
some. In both cases, however, the problem remained that the techniques,
and the software, were too convoluted and too different from what both
statisticians and users were accustomed to.

Van der Heijden and Van Buuren
(\citeproc{ref-vanderheijden_vanbuuren_16}{1916}) give an excellent,
though somewhat wistful, historic overview of the Gifi project. It is
too early for eulogies, however, and we refuse to give up. This book is
yet another reorganization of the Gifi material, with many extensions.
We take Yoshio Takane's advice seriously, and we organize both the
theory and the algorithms around what is called ``meet-loss'' in Gifi.
In our software we separate the basic computational engine from its
various applications that define the techniques of \emph{Multivariate
Analysis with Optimal Scaling (MVAOS)}. Hiding the core makes it
possible to make the programs behave in much the same way as traditional
MVA programs. The software is written in R
((\citeproc{ref-r_core_team_16}{\textbf{r\_core\_team\_16?}})), with
some parts of the computational engine written in C.

The book itself is written in Rmarkdown, using bookdown (Xie
(\citeproc{ref-xie_16}{2016})) and knitr (Xie
(\citeproc{ref-xie_15}{2015})) to embed the computations and graphics,
and to produce html and pdf versions that are completely reproducible.
The book and all the files that go with it are in the public domain.

We would like to acknowledge those who have made substantial
contributions to the Gifi project (and its immediate ancestors and
offspring) over the years. Some of them are lost in the mists of time,
some of them are no longer on this earth. They are, in alphabetical
order, Bert Bettonvil, Jason Bond, Catrien Bijleveld, Frank Busing,
Jacques Commandeur, Henny Coolen, Steef de Bie, Jan de Leeuw, John
Gower, Patrick Groenen, Chris Haveman, Willem Heiser, Abby Israels, Judy
Knip, Jan Koster, Pieter Kroonenberg, Patrick Mair, Adriaan Meester,
Jacqueline Meulman, George Michailidis, Peter Neufeglise, Dré Nierop,
Ineke Stoop, Yoshio Takane, Stef van Buuren, John van de Geer, Gerda van
den Berg, Eeke van der Burg, Peter van der Heijden, Anita van der Kooij,
Ivo van der Lans, Rien van der Leeden, Jan van Rijckevorsel, Renée
Verdegaal, Peter Verboon, Susañña Verdel, and Forrest Young.

\bookmarksetup{startatroot}

\chapter{Introduction}\label{introduction}

This is a book created from markdown and executable code.

See for additional discussion of literate programming.

\begin{Shaded}
\begin{Highlighting}[]
\DecValTok{1} \SpecialCharTok{+} \DecValTok{1}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
[1] 2
\end{verbatim}

\bookmarksetup{startatroot}

\chapter{Summary}\label{summary}

In summary, this book has no content whatsoever.

\begin{Shaded}
\begin{Highlighting}[]
\DecValTok{1} \SpecialCharTok{+} \DecValTok{1}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
[1] 2
\end{verbatim}

\bookmarksetup{startatroot}

\chapter*{References}\label{references}
\addcontentsline{toc}{chapter}{References}

\markboth{References}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-deleeuw_mair_A_09a}
De Leeuw, J., and P. Mair. 2009. {``{Homogeneity Analysis in {R}: the
Package homals}.''} \emph{Journal of Statistical Software} 31 (4):
1--21. \url{https://www.jstatsoft.org/v31/i04/}.

\bibitem[\citeproctext]{ref-gifi_B_80}
Gifi, A. 1980. \emph{Niet-Lineaire Multivariate Analyse {[}Nonlinear
Multivariate Analysis{]}}. Leiden, The Netherlands: Department of Data
Theory FSW/RUL.

\bibitem[\citeproctext]{ref-gifi_B_81}
---------. 1981. \emph{Nonlinear Multivariate Analysis}. Leiden, The
Netherlands: Department of Data Theory FSW/RUL.

\bibitem[\citeproctext]{ref-gifi_B_90}
---------. 1990. \emph{Nonlinear Multivariate Analysis}. New York, N.Y.:
Wiley.

\bibitem[\citeproctext]{ref-hill_90}
Hill, M. O. 1990. {``{Review of A. Gifi, Multivariate Analysis}.''}
\emph{Journal of Ecology} 78 (4): 1148--49.

\bibitem[\citeproctext]{ref-ibm_15}
IBM. 2015. \emph{IBM SPSS Categories 23}. IBM Corporation.

\bibitem[\citeproctext]{ref-marvell_53}
Marvell, A. 1653. {``The Character of Holland.''}

\bibitem[\citeproctext]{ref-meulman_heiser_12}
Meulman, J. J., and W. J. Heiser. 2012. \emph{IBM SPSS Categories 21}.
IBM Corporation.

\bibitem[\citeproctext]{ref-michailidis_deleeuw_A_98}
Michailidis, G., and J. De Leeuw. 1998. {``The Gifi System for
Descriptive Multivariate Analysis.''} \emph{Statistical Science} 13:
307--36.

\bibitem[\citeproctext]{ref-spss_89}
SPSS. 1989. \emph{SPSS Categories}. SPSS Inc.

\bibitem[\citeproctext]{ref-takane_92}
Takane, Y. 1992. {``{Review of Albert Gifi, Nonlinear Multivariate
Analysis}.''} \emph{Journal of the American Statistical Association} 87:
587--88.

\bibitem[\citeproctext]{ref-vanderheijden_vanbuuren_16}
Van der Heijden, P. G. M., and S. Van Buuren. 1916. {``Looking Back at
the Gifi System of Nonlinear Multivariate Analysis.''} \emph{Journal of
Statistical Software} 73 (4).

\bibitem[\citeproctext]{ref-xie_15}
Xie, Y. 2015. \emph{{Dynamic Documents with R and knitr}}. Second
Edition. CRC Press.

\bibitem[\citeproctext]{ref-xie_16}
---------. 2016. \emph{Bookdown: Authoring Books with r Markdown.}

\end{CSLReferences}




\end{document}
